Scalable gradients of Gaussian Process models with celerite

### 2018-01-25 paper/ms.tex
words

Gaussian Processes \citep[GPs][]{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some example include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is of central importance and this can often be the computational bottleneck.
The details of this model are omitted here (more details can be found in
\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
is that, for a dataset with $N$ data points, every evaluation of a GP model
requires computation of the inverse\footnote{Or, more specifically, a linear
operator that can left multiply a vector or a matrix by the inverse of this
matrix.} and log-determinant of the $N \times N$ covariance matrix.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Foreman-Mackey:2017}.
The details of these models can be found elsewhere
\citep{Foreman-Mackey:2017}, but the basic idea is that for some covariance
functions, the matrix $K$ can be written as a rank-$J$ semi-separable matrix
\begin{eqnarray}\eqlabel{semi-K}
K &=& \mathrm{diag}(\bvec{a}) + \mathrm{tril}(U\,{V}^\T) +
    \mathrm{triu}({V}\,{U}^\T)
\end{eqnarray}
where $\bvec{a}$ is a $N$-vector, ${U}$ and ${V}$ are $N \times J$
matrices, the diag function creates a diagonal matrix from a vector, and the
tril and triu extract the strict lower and upper triangular matrices from a
dense matrix.
For a matrix of this form, \citet{Foreman-Mackey:2017} derived an algorithm
with $\mathcal{O}(N\,J^2)$ scaling for computing the Cholesky factorization
\begin{eqnarray}
K &=& L\,\mathrm{diag}(\bvec{d})\,L^\T
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{semi-L}
L &=& I + \mathrm{tril}(U\,W^\T)
\end{eqnarray}
for some $N \times J$ matrix $W$.
The algorithm from \citet{Foreman-Mackey:2017} includes an $N-1 \times J$
``preconditioning'' matrix\footnote{This matrix is called $\phi$ by
\citet{Foreman-Mackey:2017}.} that we include with the notation $P$.
The preconditioned variables were indicated with a tilde by
\citet{Foreman-Mackey:2017}, but to simplify the notation, we will only refer
to the pre-conditioned variables in all that follows.

The Cholesky factorization algorithm derived by \citet{Foreman-Mackey:2017} is
given is reproduced in the following algorithm:

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_factor}($U$, $P$, $\bvec{d}$, $W$) \\
\hspace*{2em}\textsf{\# If at input $\bvec{d} = \bvec{a}$, $W = V$, and $K$ is
    given by \Eq{semi-K},} \\
\hspace*{2em}\textsf{\# at output $K = L\,L^\T$ where $L$ is given by
    \Eq{semi-L}.} \\
\hspace*{2em}$S \gets$ \texttt{zeros}($J$, $J$) \\
    \hspace*{2em}$\bvec{w}_1 \gets \bvec{w}_1 / d_{1}$ \\
\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$S \gets \texttt{diag}(\bvec{p}_{n-1})\,[
    S + d_{n-1}\,{\bvec{w}_{n-1}}^\T\,{\bvec{w}_{n-1}}
]\,\texttt{diag}(\bvec{p}_{n-1})$ \\
\hspace*{2em}\hspace*{2em}$d_{n} \gets d_{n} - \bvec{u}_n\,S\,{\bvec{u}_n}^\T$\\
\hspace*{2em}\hspace*{2em}$\bvec{w}_n \gets \left[\bvec{w}_n -
    \bvec{u}_n\,S \right] / d_{n}$\\
    \hspace*{2em}\textbf{return} $\bvec{d}$, $W$, $S$
\end{minipage}
\medskip

\noindent In this algorithm, the \texttt{zeros}$(J,\,K)$ function creates a $J
\times K$ matrix of zeros, the \texttt{diag} function creates a diagonal
matrix from a vector, and $\bvec{x}_n$ indicates a row vector made from the
$n$-th row of the matrix $X$.
Using this factorization, the log-determinant of $K$ is
\begin{eqnarray}
\log \det K &=& \sum_{n=1}^N \log d_{n} \quad.
\end{eqnarray}

Similarly, \citet{Foreman-Mackey:2017} derived a $\mathcal{O}(N\,J)$ algorithm
to apply the inverse of $K$ as shown in the following algorithm:

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_solve}($U$, $P$, $\bvec{d}$, $W$, $Z$) \\
\hspace*{2em}\textsf{\# If at input $Z = Y$, at output $Z = K^{-1}\,Y$.} \\
\hspace*{2em}$F \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
\hspace*{2em}$G \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$F \gets \texttt{diag}(\bvec{p}_{n-1})\,[F +
    {\bvec{w}_{n-1}}^\T\,\bvec{z}_{n-1}]$\\
\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n - \bvec{u}_n\,F$\\
\hspace*{2em}\textbf{for} $n = 1,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n / d_{n}$\\
\hspace*{2em}\textbf{for} $n = N-1,\ldots,1$:\\
\hspace*{2em}\hspace*{2em}$G \gets \texttt{diag}(\bvec{p}_{n})\,[G +
    {\bvec{u}_{n+1}}^\T\,\bvec{z}_{n+1}]$\\
    \hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n - \bvec{w}_n\,G$\\
\hspace*{2em}\textbf{return} $Z$, $F$, $G$
\end{minipage}
\medskip

@@

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_factor\_grad}($U$, $P$, $\bvec{d}$, $W$,
    $S$, $\bar{S}$, $\bar{\bvec{a}}$, $\bar{V}$) \\
\hspace*{2em}\textsf{\# At input $\bar{\bvec{a}} = \bar{\bvec{d}}$ and
    $\bar{V} = \bar{W}$}\\
    \hspace*{2em}$\bar{\bvec{v}}_N \gets \bar{\bvec{v}}_N / d_N$\\
\hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
\hspace*{2em}\hspace*{2em}$\bar{a}_n \gets \bar{a}_n -
    \bvec{w}_n\,{\bar{\bvec{v}}_n}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_n \gets - [\bar{\bvec{v}}_n +
    2\,\bar{a}_n\,\bvec{u}_n]\,S$\\
\hspace*{2em}\hspace*{2em}$\bar{S} \gets \bar{S} -
    {\bvec{u}_n}^\T\,[\bar{\bvec{v}}_n + \bar{a}_n\,\bvec{u}_n]$\\

\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_{n-1} \gets \mathrm{diag}(\bar{S}\,S
    \,\mathrm{diag}(\bvec{p}_{n-1})^{-1} + \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,
    S\,\bar{S})$\\
\hspace*{2em}\hspace*{2em}$\bar{S} \gets \mathrm{diag}(\bvec{p}_{n-1})\,
    \bar{S}\,\mathrm{diag}(\bvec{p}_{n-1})$\\
\hspace*{2em}\hspace*{2em}$\bar{d}_{n-1} \gets \bar{d}_{n-1} +
    \bvec{w}_{n-1}\,\bar{S}\,{\bvec{w}_{n-1}}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{v}}_{n-1} \gets \bar{\bvec{v}}_{n-1}
    / d_{n-1} + \bvec{w}_{n-1}\,[\bar{S} + \bar{S}^\T]$\\

\hspace*{2em}\hspace*{2em}$S \gets \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,S\,
    \mathrm{diag}(\bvec{p}_{n-1})^{-1}
    - d_{n-1}\,{\bvec{w}_{n-1}}^\T\,\bvec{w}_{n-1}$\\

\hspace*{2em}$\bar{a}_1 \gets \bar{a}_1 -
    \bar{\bvec{v}}_1\,{\bvec{w}_1}^\T$\\

\hspace*{2em}\textbf{return} $\bar{U}$, $\bar{P}$, $\bar{\bvec{a}}$, $\bar{V}$
\end{minipage}
\medskip

>>>>

Gaussian Processes \citep[GPs][]{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some example include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is of central importance and this can often be the computational bottleneck.
The details of this model are omitted here (more details can be found in
\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
is that, for a dataset with $N$ data points, every evaluation of a GP model
requires computation of the inverse\footnote{Or, more specifically, a linear
operator that can left multiply a vector or a matrix by the inverse of this
matrix.} and log-determinant of the $N \times N$ covariance matrix.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Foreman-Mackey:2017}.

The details of the \emph{celerite} method can be found in
\citet{Foreman-Mackey:2017} and I will not repeat them here.
The only difference in notation here is that all the matrices to which I refer
are the ``pre-conditioned'' matrices that are indicated with a tilde by
\citet{Foreman-Mackey:2017}, but the tilde is not included here for
simplicity.
I will also use the symbol $P$ to mean the $N-1 \times J$ pre-conditioning
matrix that was called $\phi$ by \citet{Foreman-Mackey:2017}.
The Cholesky factorization algorithm derived by \citet[][their
Equation~46]{Foreman-Mackey:2017} is given is listed below in the following
algorithm:

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_factor}($U$, $P$, $\bvec{d}$, $W$) \\
\hspace*{2em}\textsf{\# At input $\bvec{d} = \bvec{a}$ and $W = V$} \\
\hspace*{2em}$S \gets$ \texttt{zeros}($J$, $J$) \\
    \hspace*{2em}$\bvec{w}_1 \gets \bvec{w}_1 / d_{1}$ \\
\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$S \gets \texttt{diag}(\bvec{p}_{n-1})\,[
    S + d_{n-1}\,{\bvec{w}_{n-1}}^\T\,{\bvec{w}_{n-1}}
]\,\texttt{diag}(\bvec{p}_{n-1})$ \\
\hspace*{2em}\hspace*{2em}$d_{n} \gets d_{n} - \bvec{u}_n\,S\,{\bvec{u}_n}^\T$\\
\hspace*{2em}\hspace*{2em}$\bvec{w}_n \gets \left[\bvec{w}_n -
    \bvec{u}_n\,S \right] / d_{n}$\\
    \hspace*{2em}\textbf{return} $\bvec{d}$, $W$, $S$
\end{minipage}
\medskip

\noindent In this algorithm, the \texttt{zeros}$(J,\,K)$ function creates a $J
\times K$ matrix of zeros, the \texttt{diag} function creates a diagonal
matrix from a vector, and $\bvec{x}_n$ indicates a row vector made from the
$n$-th row of the matrix $X$.
Using this factorization, the log-determinant of $K$ is
\begin{eqnarray}
\log \det K &=& \sum_{n=1}^N \log d_{n} \quad.
\end{eqnarray}
Similarly, \citet{Foreman-Mackey:2017} derived a $\mathcal{O}(N\,J)$ algorithm
to apply the inverse of $K$ as follows (Equations~47 and~48 in
\citealt{Foreman-Mackey:2017}):

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_solve}($U$, $P$, $\bvec{d}$, $W$, $Z$) \\
\hspace*{2em}\textsf{\# At input $Z = Y$} \\
\hspace*{2em}$F \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
\hspace*{2em}\textbf{for} $n = 2,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$F \gets \texttt{diag}(\bvec{p}_{n-1})\,[F +
    {\bvec{w}_{n-1}}^\T\,\bvec{z}_{n-1}]$\\
\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n - \bvec{u}_n\,F$\\
\hspace*{2em}\textbf{for} $n = 1,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n / d_{n}$\\
\hspace*{2em}$G \gets$ \texttt{zeros}($J$, $N_\mathrm{rhs}$) \\
\hspace*{2em}\textbf{for} $n = N-1,\ldots,1$:\\
\hspace*{2em}\hspace*{2em}$G \gets \texttt{diag}(\bvec{p}_{n})\,[G +
    {\bvec{u}_{n+1}}^\T\,\bvec{z}_{n+1}]$\\
    \hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n - \bvec{w}_n\,G$\\
\hspace*{2em}\textbf{return} $Z$, $F$, $G$
\end{minipage}
\medskip

@@

Using the notation from \citet{Giles:2008} and performing some tedious
algebra, the reverse-mode implementation of the \texttt{celerite\_factor}
function is found to be:

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_factor\_rev}($U$, $P$, $\bvec{d}$, $W$,
    $S$, $\bar{S}$, $\bar{\bvec{a}}$, $\bar{V}$) \\
\hspace*{2em}\textsf{\# At input $\bar{\bvec{a}} = \bar{\bvec{d}}$ and
    $\bar{V} = \bar{W}$}\\
    \hspace*{2em}$\bar{\bvec{v}}_N \gets \bar{\bvec{v}}_N / d_N$\\
\hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
\hspace*{2em}\hspace*{2em}$\bar{a}_n \gets \bar{a}_n -
    \bvec{w}_n\,{\bar{\bvec{v}}_n}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_n \gets - [\bar{\bvec{v}}_n +
    2\,\bar{a}_n\,\bvec{u}_n]\,S$\\
\hspace*{2em}\hspace*{2em}$\bar{S} \gets \bar{S} -
    {\bvec{u}_n}^\T\,[\bar{\bvec{v}}_n + \bar{a}_n\,\bvec{u}_n]$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_{n-1} \gets \mathrm{diag}(\bar{S}\,S
    \,\mathrm{diag}(\bvec{p}_{n-1})^{-1} + \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,
    S\,\bar{S})$\\
\hspace*{2em}\hspace*{2em}$\bar{S} \gets \mathrm{diag}(\bvec{p}_{n-1})\,
    \bar{S}\,\mathrm{diag}(\bvec{p}_{n-1})$\\
\hspace*{2em}\hspace*{2em}$\bar{d}_{n-1} \gets \bar{d}_{n-1} +
    \bvec{w}_{n-1}\,\bar{S}\,{\bvec{w}_{n-1}}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{v}}_{n-1} \gets \bar{\bvec{v}}_{n-1}
    / d_{n-1} + \bvec{w}_{n-1}\,[\bar{S} + \bar{S}^\T]$\\
\hspace*{2em}\hspace*{2em}$S \gets \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,S\,
    \mathrm{diag}(\bvec{p}_{n-1})^{-1}
    - d_{n-1}\,{\bvec{w}_{n-1}}^\T\,\bvec{w}_{n-1}$\\
\hspace*{2em}$\bar{a}_1 \gets \bar{a}_1 -
    \bar{\bvec{v}}_1\,{\bvec{w}_1}^\T$\\
\hspace*{2em}\textbf{return} $\bar{U}$, $\bar{P}$, $\bar{\bvec{a}}$, $\bar{V}$
\end{minipage}
\medskip

\noindent Similarly, the reverse-mode function corresponding to
\texttt{celerite\_solve} is:

\medskip
\begin{minipage}{\linewidth}
\textbf{function} \texttt{celerite\_solve\_rev}($U$, $P$, $\bvec{d}$, $W$,
    $Z$, $F$, $G$, $\bar{F}$, $\bar{G}$, $\bar{Y}$) \\
\hspace*{2em}\textsf{\# At input $\bar{Y} = \bar{Z}$}\\
\hspace*{2em}\textbf{for} $n = 1,\ldots,N-1$:\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{w}}_n \gets
    - \bar{\bvec{y}}_n\,G^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{G} \gets \bar{G} - {\bvec{w}_n}^\T \,
    \bar{\bvec{y}}_n$\\
\hspace*{2em}\hspace*{2em}$\bvec{z}_n \gets \bvec{z}_n + \bvec{w}_n\,G$\\
\hspace*{2em}\hspace*{2em}$G \gets \mathrm{diag}(\bvec{p}_n)^{-1}\,G$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_n \gets \mathrm{diag}(
    \bar{G}\,G^\T)$\\
\hspace*{2em}\hspace*{2em}$\bar{G} \gets
    \mathrm{diag}(\bvec{p}_n)\,\bar{G}$\\
\hspace*{2em}\hspace*{2em}$G \gets G - {\bvec{u}_{n+1}}^\T\,\bvec{z}_{n+1}$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_{n+1} \gets
    \bvec{z}_{n+1}\,\bar{G}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{y}}_{n+1} \gets
    \bvec{u}_{n+1}\,\bar{G}$\\
\hspace*{2em}\textbf{for} $n = 1,\ldots,N$:\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{y}}_n \gets \bar{\bvec{y}}_n/d_n$\\
\hspace*{2em}\hspace*{2em}$\bar{d}_n \gets -
    \bvec{z}_n\,{\bar{\bvec{y}}_n}^\T$\\
\hspace*{2em}\textbf{for} $n = N,\ldots,2$:\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{u}}_n \gets \bar{\bvec{u}}_n
    - \bar{\bvec{y}}_n\,F^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{F} \gets \bar{F} - {\bvec{u}_n}^\T \,
    \bar{\bvec{y}}_n$\\
\hspace*{2em}\hspace*{2em}$F \gets \mathrm{diag}(\bvec{p}_{n-1})^{-1}\,F$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{p}}_{n-1} \gets \bar{\bvec{p}}_{n-1} +
    \mathrm{diag}(\bar{F}\,F^\T)$\\
\hspace*{2em}\hspace*{2em}$\bar{F} \gets
    \mathrm{diag}(\bvec{p}_{n-1})\,\bar{F}$\\
\hspace*{2em}\hspace*{2em}$F \gets F - {\bvec{w}_{n-1}}^\T\,\bvec{z}_{n-1}$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{w}}_{n-1} \gets \bar{\bvec{w}}_{n-1} +
    \bvec{z}_{n-1}\,\bar{F}^\T$\\
\hspace*{2em}\hspace*{2em}$\bar{\bvec{y}}_{n-1} \gets \bar{\bvec{y}}_{n-1} +
    \bvec{w}_{n-1}\,\bar{F}$\\
\hspace*{2em}\textbf{return} $\bar{U}$, $\bar{P}$, $\bar{\bvec{d}}$,
    $\bar{W}$, $\bar{Y}$
\end{minipage}
\medskip

\noindent A reference C++ implementation of this algorithm can be found online
TODO ADD ZENODO LINK and \Figure{figure} shows the performance of this
implementation.

\section{Discussion}

This research note presents the algorithms needed to efficiently compute
gradients of GP models applied to large datasets using the \emph{celerite}
method.
These developments increase the performance of inference methods based on
\emph{celerite} and improve the convergence properties of non-linear
optimization routines.
Furthermore, the derivation of reverse-mode accumulation algorithms for
\emph{celerite} allow its integration into popular model building and
automatic differentiation libraries like Stan \citep{Carpenter:2015},
TensorFlow \citep{Abadi:2016}, and others.
